---
title: "Divvy Case Study, Data Cleaning Report"
author: "Tanner Quesenberry"
date: "7/13/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load-packages, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(magrittr)
library(knitr)
```

## Data Cleaning Steps

* Install and load the `tidyverse`, `lubridate`, `ggplot2` and `dplyr` packages.
* Read in the .csv data files for the months of June 2020 to May 2021.
``` {r echo=FALSE, message=FALSE, results='hide'}
jun_2020_to_may_2021 <- read_csv("divvy-cleaned-data.csv")
```
  
* Join the tables from June 2020 to Nov 2020 together with `full_join()` to get combined data frames.
Example `apr_may <- full_join(apr_2021_trips, may_2021_trips)`
* Repeat table joining for tables from Dec 2020 to May 2021.
* Resolve the incompatible type difference of `start_station_id` and `end_station_id` columns by casting to characters. `jun_nov <- jun_nov %>% mutate(start_station_id = as.character(start_station_id))`
`jun_nov <- jun_nov %>% mutate(end_station_id = as.character(end_station_id))`
* Join these two 6 month tables to form a single one year table for analysis.
* Save the table to a .csv file for later reference if needed.
`write.csv(jun_2020_to_may_2021, "~/Desktop/Programming/DataAnalyst/Capstone/202006-202105-divvy-tripdata.csv", row.names = FALSE)`  

* Found 209 duplicate rows exist using 
`jun_2020_to_may_2021 %>% summarise(count = n_distinct(ride_id))`
* Removed duplicate rows `jun_2020_to_may_2021 <- jun_2020_to_may_2021 %>% distinct(ride_id, .keep_all = TRUE)`
* Verified that member_casual contains correct data
`View(filter(jun_2020_to_may_2021, member_casual != "casual", member_casual != "member"))`
* Verified `start_at` and `ended_at` had times for all rows through viewing and sorting the table.
* Checked that `rideable_type` column contained the correct bike options.
`jun_2020_to_may_2021 %>% distinct(rideable_type)`


* Removed the `start_station_id` and `end_station_id` as they serve no purpose in this analysis.
`jun_2020_to_may_2021 <- subset(jun_2020_to_may_2021, select = -c(start_station_id,end_station_id))`
* Similarly removed `start_lat`, `start_lng`, `end_lat`, `end_lng`
`jun_2020_to_may_2021 <- subset(jun_2020_to_may_2021, select = -c(start_lat,start_lng,end_lat,end_lng))`


* Created an additional column `day_of_week` to indicate which weekday the ride starts on.
`jun_2020_to_may_2021 <- jun_2020_to_may_2021 %>% mutate(day_of_week = wday(started_at))`
* Created additional columns for date, month, day, and year to aggregate on later.
`jun_2020_to_may_2021$date <- as.Date(jun_2020_to_may_2021$started_at)`
`jun_2020_to_may_2021$month <- format(as.Date(jun_2020_to_may_2021$date), "%m")`
`jun_2020_to_may_2021$day <- format(as.Date(jun_2020_to_may_2021$date), "%d")`
`jun_2020_to_may_2021$year <- format(as.Date(jun_2020_to_may_2021$date), "%Y")`
* Created an additional column for `ride_length_secs`. Then removed any row with a negative duration, and converted to a time format column `ride_length`.
`jun_2020_to_may_2021 <- mutate(jun_2020_to_may_2021, ride_length_secs = (ended_at - started_at))`
`jun_2020_to_may_2021 <- subset(jun_2020_to_may_2021, jun_2020_to_may_2021$ride_length_secs > 0)`
`jun_2020_to_may_2021 <- mutate(jun_2020_to_may_2021, ride_length = hms::hms(seconds_to_period(jun_2020_to_may_2021$ride_length_secs)))`


* Saved the cleaned data set as divvy-cleaned-data.csv
`write.csv(jun_2020_to_may_2021, "~/Desktop/Programming/DataAnalyst/Capstone/divvy-cleaned-data.csv", row.names = FALSE)`


## Descriptive Analysis

##### Gathering the descriptive statistics for the user ride lengths.

``` {r echo=FALSE, results='hold', comment="" }
cat("Ride length (seconds)", "\n")

cat("Min: ", min(jun_2020_to_may_2021$ride_length_secs), "\n")
cat("Max: ", max(jun_2020_to_may_2021$ride_length_secs), "\n")
cat("Median: ", median(jun_2020_to_may_2021$ride_length_secs), "\n")
cat("Mean: ", mean(jun_2020_to_may_2021$ride_length_secs), "\n")
```

##### Comparing the ride lengths of members vs casual riders

``` {r echo=FALSE, results='hold', comment="" }
Rider <- jun_2020_to_may_2021$member_casual
length_secs <- jun_2020_to_may_2021$ride_length_secs

cat("Mean\n")
aggregate(length_secs ~ Rider, FUN = mean)
cat("\nMedian\n")
aggregate(length_secs ~ Rider, FUN = median)
cat("\nMax\n")
aggregate(length_secs ~ Rider, FUN = max)
cat("\nMin\n")
aggregate(length_secs ~ Rider, FUN = min)

```

##### Comparing casual vs member ride lengths by day of week

``` {r echo=FALSE, comment=""}
Rider <- jun_2020_to_may_2021$member_casual
length_secs <- jun_2020_to_may_2021$ride_length_secs
weekday <- jun_2020_to_may_2021$day_of_week
cat("Sunday = 1 ... Saturday = 7\n")
aggregate(length_secs ~ Rider + weekday, FUN = mean)
```


##### Analyze ridership by type and weekday

``` {r echo=FALSE, message=FALSE}
jun_2020_to_may_2021 %>% mutate(weekday = wday(started_at, label = TRUE)) %>% group_by(member_casual, weekday) %>% summarise(number_of_rides = n(), average_duration = mean(ride_length_secs)) %>% arrange(member_casual, weekday)
```



## Visualize ridership by type

```{r pressure, echo=FALSE, message=FALSE}
options(scipen=5)
jun_2020_to_may_2021 %>% 
    mutate(weekday = wday(started_at, label = TRUE)) %>% 
    group_by(member_casual, weekday) %>% 
    summarise(number_of_rides = n()
              ,average_duration = mean(ride_length_secs)) %>% 
    arrange(member_casual, weekday)  %>% 
    ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
    geom_col(position = "dodge") + ylab("Total Rides") + xlab("Day of Week") +ggtitle("Total Rides Per Weekday")
```

```{r echo=FALSE, message=FALSE}
jun_2020_to_may_2021 %>% 
    mutate(weekday = wday(started_at, label = TRUE)) %>% 
    group_by(member_casual, weekday) %>% 
    summarise(number_of_rides = n()
              ,average_duration = (mean(ride_length_secs)/60)) %>% 
    arrange(member_casual, weekday)  %>% 
    ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
    geom_col(position = "dodge") + xlab("Day of Week") + ylab("Average Ride Length (Minutes)") + ggtitle("Average Ride Time By Day")
```